---
title: "Data visualization with ggplot2"
author: "David Gresham"
date: "Last compiled on `r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
---

## Introduction

`ggplot2` is R's most widely used plotting library, built around a principled framework called the **Grammar of Graphics** (Wilkinson, 2005). The core idea is that every plot — no matter how complex — can be described by the same set of components. Once you internalize this grammar, you can build any plot you can imagine by combining these components, rather than hunting for a specific function (like `barplot()`, `hist()`, `boxplot()`) for each plot type.

This contrasts with base R plotting, where each chart type has its own syntax and customization options that you have to learn separately. In `ggplot2`, all plots share the same underlying structure.

## The Grammar of Graphics

A ggplot2 plot is built from the following layers:

| Layer | Function(s) | What it controls |
|---|---|---|
| **Data** | `ggplot(data = ...)` | The dataframe to plot |
| **Aesthetics** | `aes(x, y, color, fill, ...)` | How variables map to visual properties |
| **Geometries** | `geom_point()`, `geom_histogram()`, ... | The type of mark drawn on the plot |
| **Statistics** | `stat_summary()`, `geom_smooth()`, ... | Statistical transformations of the data |
| **Scales** | `scale_x_log10()`, `scale_color_brewer()`, ... | How aesthetic mappings are rendered |
| **Coordinates** | `coord_flip()`, `coord_cartesian()` | The coordinate space |
| **Facets** | `facet_wrap()`, `facet_grid()` | Splitting into subplots by a variable |
| **Themes** | `theme_bw()`, `theme()`, ... | Non-data visual elements (fonts, grid, background) |

You build a plot by starting with `ggplot()` and adding layers with `+`. This makes it easy to build up complexity incrementally and to understand exactly what each piece of code is doing.

## Setup

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggthemes)   # extra themes; install with install.packages("ggthemes") if needed
```

## Example dataset: *S. cerevisiae* genome features

Throughout this vignette we use the *S. cerevisiae* R64 genome annotation (GFF3 format) as our dataset. This gives us a real biological dataset with a mix of continuous (feature length, chromosomal position) and categorical (chromosome, feature type, strand) variables — a common situation in genomics.

We load the GFF3 and derive a clean dataframe of genome features with their lengths.

```{r message=FALSE, warning=FALSE}
# Read the GFF3, skipping the header comment lines
gff <- read_delim("../data/Saccharomyces_cerevisiae.R64-1-1.34.gff3",
    delim = "\t", col_names = FALSE,
    comment = "#", trim_ws = TRUE, skip = 24)

names(gff) <- c("chromosome", "source", "feature",
                "start", "stop", "score", "strand", "phase", "info")

# Set categorical columns as factors
gff$feature    <- as.factor(gff$feature)
gff$chromosome <- as.factor(gff$chromosome)
gff$strand     <- as.factor(gff$strand)

# Subset to feature types of interest; calculate length
yeast_features <- gff %>%
    select(chromosome, feature, start, stop, strand) %>%
    mutate(length = abs(stop - start)) %>%
    filter(feature %in% c("CDS", "rRNA", "snoRNA", "snRNA", "tRNA_gene"))

head(yeast_features)
```

The dataset has `r nrow(yeast_features)` rows — one per annotated genomic feature — and covers `r nlevels(yeast_features$chromosome)` chromosomes and `r nlevels(yeast_features$feature)` feature types.

```{r}
# Feature types and chromosomes present
levels(yeast_features$feature)
levels(yeast_features$chromosome)
```

---

## Building a plot layer by layer

The key habit to develop with ggplot2 is building plots incrementally. Start with the data and aesthetics, add a geometry, then refine.

### Step 1: data + aesthetics

`ggplot()` alone produces an empty canvas — it knows what data to use and which variables map to x and y, but hasn't been told how to draw anything yet.

```{r}
ggplot(data = yeast_features, mapping = aes(x = chromosome, y = length))
```

### Step 2: add a geometry

Add `geom_boxplot()` to draw the data.

```{r warning=FALSE}
ggplot(data = yeast_features, mapping = aes(x = chromosome, y = length)) +
    geom_boxplot()
```

### Step 3: add more layers

Layers accumulate. Add a log scale and jittered raw data points on top of the boxplot.

```{r warning=FALSE}
ggplot(data = yeast_features, mapping = aes(x = chromosome, y = length)) +
    geom_boxplot() +
    geom_jitter(aes(color = feature), alpha = 0.2, size = 0.5) +
    scale_y_log10()
```

This pattern — start simple, add layers — is the core ggplot2 workflow.

---

## Aesthetics

Aesthetics (`aes()`) map variables to visual properties. The most common are:

- `x`, `y` — position
- `color` / `colour` — outline or line color
- `fill` — interior color (for bars, boxplots, histograms)
- `alpha` — transparency (0 = invisible, 1 = opaque)
- `size` — point size or line width
- `shape` — point shape
- `linetype` — line type

**Important distinction:** aesthetics set *inside* `aes()` map a variable to a visual property; aesthetics set *outside* `aes()` apply a fixed value to all observations.

```{r warning=FALSE}
# color mapped to feature type (inside aes) — different color per feature
ggplot(data = yeast_features, mapping = aes(x = chromosome, y = length)) +
    geom_jitter(aes(color = feature), alpha = 0.3, size = 0.5)
```

```{r warning=FALSE}
# color fixed to "steelblue" (outside aes) — all points the same color
ggplot(data = yeast_features, mapping = aes(x = chromosome, y = length)) +
    geom_jitter(color = "steelblue", alpha = 0.3, size = 0.5)
```

---

## Common geometries

### Distributions: histograms and frequency polygons

```{r message=FALSE, warning=FALSE}
# Histogram of feature lengths
ggplot(data = yeast_features, mapping = aes(x = length)) +
    geom_histogram()
```

CDS lengths are highly right-skewed — a log scale is more informative.

```{r message=FALSE, warning=FALSE}
ggplot(data = yeast_features, mapping = aes(x = length)) +
    geom_histogram() +
    scale_x_log10()
```

`geom_freqpoly()` draws the same data as a line, which makes it easier to compare distributions across groups.

```{r message=FALSE, warning=FALSE}
# Compare distributions across feature types
ggplot(data = yeast_features, mapping = aes(x = length, color = feature)) +
    geom_freqpoly() +
    scale_x_log10()
```

To compare groups with different counts, plot as a probability density using `after_stat(density)`.

```{r message=FALSE, warning=FALSE}
ggplot(data = yeast_features, mapping = aes(x = length,
                                             y = after_stat(density),
                                             color = feature)) +
    geom_freqpoly() +
    scale_x_log10()
```

### Scatter plots

`geom_point()` for two continuous variables. With many points, use `alpha` to reveal density.

```{r warning=FALSE}
ggplot(data = yeast_features, mapping = aes(x = start, y = length)) +
    geom_point(aes(color = feature), alpha = 0.2, size = 0.8) +
    scale_y_log10()
```

### Dealing with overplotting on categorical axes

When x is categorical, `geom_point()` stacks all points directly on top of each other. Use `geom_jitter()` to add random horizontal noise so individual points are visible.

```{r warning=FALSE}
ggplot(data = yeast_features, mapping = aes(x = chromosome, y = length)) +
    geom_jitter(aes(color = feature), alpha = 0.15, size = 0.5) +
    scale_y_log10()
```

### Boxplots and violin plots

Boxplots and violin plots summarize distributions across groups. Violin plots are often more informative because they show the full distribution shape, not just quartiles.

```{r warning=FALSE}
# Boxplot
ggplot(data = yeast_features, mapping = aes(x = chromosome, y = length)) +
    geom_boxplot() +
    scale_y_log10()
```

```{r warning=FALSE}
# Violin plot — shows distribution shape
ggplot(data = yeast_features, mapping = aes(x = feature, y = length, fill = feature)) +
    geom_violin() +
    scale_y_log10() +
    theme(legend.position = "none")   # legend is redundant when x = fill
```

Combine layers to show both summary and raw data:

```{r warning=FALSE}
ggplot(data = yeast_features, mapping = aes(x = feature, y = length)) +
    geom_violin(fill = "grey90") +
    geom_jitter(aes(color = feature), alpha = 0.2, size = 0.5) +
    scale_y_log10() +
    theme(legend.position = "none")
```

### Bar plots

`geom_bar()` counts observations in each group by default (`stat = "count"`).

```{r}
# Count features per chromosome, colored by feature type
ggplot(data = yeast_features, mapping = aes(x = chromosome, fill = feature)) +
    geom_bar()
```

The `position` argument controls how bars for multiple groups are arranged:

```{r}
# "fill" normalizes to 100% — useful for comparing proportions
ggplot(data = yeast_features, mapping = aes(x = chromosome, fill = feature)) +
    geom_bar(position = "fill") +
    labs(y = "proportion")
```

```{r}
# "dodge" places bars side by side
ggplot(data = yeast_features, mapping = aes(x = chromosome, fill = feature)) +
    geom_bar(position = "dodge")
```

---

## Scales

Scales control how aesthetic mappings are rendered — axis limits, transformations, color palettes, etc.

### Axis transformations

Log transformation is essential for biological data that spans orders of magnitude (gene expression, feature lengths, allele frequencies).

```{r warning=FALSE}
# Without log scale — skewed data is uninterpretable
ggplot(data = yeast_features, mapping = aes(x = feature, y = length)) +
    geom_boxplot()
```

```{r warning=FALSE}
# With log scale — distributions are clear
ggplot(data = yeast_features, mapping = aes(x = feature, y = length)) +
    geom_boxplot() +
    scale_y_log10()
```

### Reordering categorical axes

By default, factor levels are ordered alphabetically. Use `reorder()` to sort by a summary statistic (e.g., median), which often makes plots much easier to read.

```{r warning=FALSE}
ggplot(data = yeast_features,
       mapping = aes(x = reorder(chromosome, length, FUN = median), y = length)) +
    geom_boxplot() +
    scale_y_log10() +
    labs(x = "chromosome")
```

### Color scales

For **colorblind accessibility**, prefer `scale_color_viridis_d()` (discrete) or `scale_color_viridis_c()` (continuous) over the ggplot2 default palette.

```{r warning=FALSE}
ggplot(data = yeast_features,
       mapping = aes(x = length, y = after_stat(density), color = feature)) +
    geom_freqpoly() +
    scale_x_log10() +
    scale_color_viridis_d()
```

---

## Facets

Faceting splits the data into subplots by one or more variables. This is one of ggplot2's most powerful features for exploring multi-dimensional data.

`facet_wrap()` wraps subplots into a grid by one variable:

```{r warning=FALSE, fig.width=10}
ggplot(data = yeast_features, mapping = aes(x = length, fill = chromosome)) +
    geom_histogram(show.legend = FALSE) +
    scale_x_log10() +
    facet_wrap(~chromosome)
```

`facet_grid()` creates a two-way grid using two variables:

```{r warning=FALSE, fig.width=10}
ggplot(data = yeast_features, mapping = aes(x = length)) +
    geom_histogram() +
    scale_x_log10() +
    facet_grid(feature ~ strand)
```

---

## Statistics

Some geoms compute statistics internally (e.g., `geom_histogram()` bins data, `geom_boxplot()` computes quartiles). You can also add explicit statistical layers.

### Trend lines with `geom_smooth()`

```{r warning=FALSE}
# Default: LOESS smoother
ggplot(data = yeast_features, mapping = aes(x = start, y = stop)) +
    geom_point(aes(color = feature), alpha = 0.2, size = 0.5) +
    geom_smooth() +
    scale_x_log10() +
    scale_y_log10()
```

```{r warning=FALSE}
# Linear model
ggplot(data = yeast_features, mapping = aes(x = start, y = stop)) +
    geom_point(aes(color = feature), alpha = 0.2, size = 0.5) +
    geom_smooth(method = "lm") +
    scale_x_log10() +
    scale_y_log10()
```

### Summary statistics with `stat_summary()`

```{r}
# Mean ± 2 SD (mean_sdl)
ggplot(data = yeast_features, mapping = aes(x = feature, y = length)) +
    stat_summary(fun.data = mean_sdl) +
    scale_y_log10()
```

---

## Labels and themes

### Labels

Every published plot should have axis labels and a title. Use `labs()`:

```{r message=FALSE, warning=FALSE}
ggplot(data = yeast_features,
       mapping = aes(x = length, y = after_stat(density), color = feature)) +
    geom_freqpoly() +
    scale_x_log10() +
    scale_color_viridis_d() +
    labs(
        title = "Size distribution of S. cerevisiae genome features",
        x = "Feature length (bp)",
        y = "Probability density",
        color = "Feature type"
    )
```

### Themes

Themes control all non-data visual elements. Built-in options:

```{r message=FALSE, warning=FALSE}
p <- ggplot(data = yeast_features,
            mapping = aes(x = length, y = after_stat(density), color = feature)) +
    geom_freqpoly() +
    scale_x_log10() +
    scale_color_viridis_d() +
    labs(x = "Feature length (bp)", y = "Probability density", color = "Feature type")

p + theme_bw()       # clean white background — good for presentations
p + theme_classic()  # minimal — good for publications
p + theme_light()    # light grey gridlines
```

`ggthemes` provides additional options including Tufte's minimalist style:

```{r message=FALSE, warning=FALSE}
p + theme_tufte()
```

You can fine-tune any element within a theme using `theme()`. A common adjustment is moving the legend:

```{r message=FALSE, warning=FALSE}
p + theme_classic() + theme(legend.position = "bottom")
```

---

## Saving plots

Save a plot as a variable and build on it incrementally. When you're done, use `ggsave()` to write a publication-quality file.

```{r}
# Build and save to a variable
my_plot <- ggplot(data = yeast_features,
                  mapping = aes(x = feature, y = length, fill = feature)) +
    geom_violin() +
    geom_jitter(alpha = 0.1, size = 0.4) +
    scale_y_log10() +
    scale_fill_viridis_d() +
    labs(
        title = "Feature length distributions in S. cerevisiae",
        x = "Feature type",
        y = "Length (bp)"
    ) +
    theme_classic() +
    theme(legend.position = "none")

my_plot
```

```{r eval=FALSE}
# Save as PDF (vector format — preferred for publication) or PNG
ggsave("feature_lengths.pdf", plot = my_plot, width = 6, height = 4)
ggsave("feature_lengths.png", plot = my_plot, width = 6, height = 4, dpi = 300)
```

Storing a plot as a variable also lets you add layers to a finished plot:

```{r warning=FALSE}
# Add a horizontal reference line to an existing plot
my_plot + geom_hline(yintercept = 1000, color = "red", linetype = "dashed")
```

---

## Coordinate systems

`coord_flip()` swaps x and y axes — useful when category labels are long:

```{r warning=FALSE}
ggplot(data = yeast_features,
       mapping = aes(x = reorder(chromosome, length, FUN = median), y = length)) +
    geom_boxplot() +
    scale_y_log10() +
    labs(x = "chromosome") +
    coord_flip()
```

`coord_cartesian()` zooms without dropping data (unlike setting axis limits in `scale_*`):

```{r warning=FALSE}
ggplot(data = yeast_features, mapping = aes(x = start, y = stop)) +
    geom_point(aes(color = feature), alpha = 0.2, size = 0.5) +
    geom_smooth() +
    coord_cartesian(xlim = c(0, 500000), ylim = c(0, 500000))
```

---

## Best practices

**Use `theme_classic()` or `theme_bw()` for publication figures.** The default grey background looks fine on screen but prints poorly and is rarely appropriate for papers.

**Label all axes.** Always include units (bp, nt, RPKM, etc.). Use `labs()`.

**Use colorblind-safe palettes.** `scale_color_viridis_d()` and `scale_fill_viridis_d()` are built into ggplot2, perceptually uniform, and work in greyscale.

**Log-transform skewed biological data.** Expression values, feature lengths, allele frequencies — most biological measurements are log-normally distributed. `scale_y_log10()` (or `scale_x_log10()`) makes these distributions readable without modifying the underlying data.

**Show the data.** Overlay raw points on boxplots or violin plots with `geom_jitter()`. Summary statistics alone obscure the distribution and sample size.

**Save plots to variables.** This makes it easy to iterate, add layers, or export at different sizes.

**Use `ggsave()` for output.** It respects the `width`, `height`, and `dpi` arguments. For publication, use PDF or SVG (vector formats); for web/slides, use PNG at 150–300 dpi.

---

## The full syntax

Any ggplot2 call follows this template (most layers are optional):

```r
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +
    <GEOM_FUNCTION>(aes(<OPTIONAL_MAPPINGS>)) +
    <STAT_FUNCTION>() +
    <SCALE_FUNCTION>() +
    <FACET_FUNCTION>() +
    <COORD_FUNCTION>() +
    <THEME_FUNCTION>() +
    labs(<LABELS>)
```

---

## Exercises

### Exercise 1
Using `yeast_features`, create a histogram of feature lengths, faceted by `feature` type. Use a log10 x-axis and `theme_classic()`. Add appropriate axis labels.

```{r message=FALSE, warning=FALSE}
# Your code here
ggplot(data = yeast_features, mapping = aes(x = length)) +
    geom_histogram()
```

### Exercise 2
Recreate the bar plot of feature counts per chromosome, but with bars **dodged** (side by side). Then switch `position = "fill"` to show proportions instead of counts. Which view is more informative for comparing the relative composition of chromosomes?

```{r}
# Your code here
ggplot(data = yeast_features, mapping = aes(x = chromosome, fill = feature)) +
    geom_bar()
```

### Exercise 3
Make a scatter plot of `start` vs. `length` for CDS features only (filter first with `dplyr::filter()`). Color points by `chromosome`, use a log10 y-axis, add a linear trend line with `geom_smooth(method = "lm")`, and apply `theme_bw()`.

```{r warning=FALSE}
# Your code here
cds_only <- yeast_features %>% filter(feature == "CDS")
ggplot(data = cds_only, mapping = aes(x = start, y = length))
```
