---
title: "Data wrangling with the tidyverse"
author: "David Gresham"
date: "Last compiled on `r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
---

## Introduction

The **tidyverse** is a collection of R packages that share a common philosophy and syntax for working with data. The core packages are loaded together with `library(tidyverse)`:

| Package | Purpose |
|---|---|
| `readr` | Reading rectangular data from files |
| `dplyr` | Data manipulation (filter, select, summarize, join) |
| `tidyr` | Reshaping data between wide and long formats |
| `stringr` | String manipulation |
| `ggplot2` | Data visualization |
| `forcats` | Working with factors |
| `purrr` | Functional programming (applying functions across lists) |
| `tibble` | A modern reimagining of the data frame |

The tidyverse is built around the concept of **tidy data**, which makes all of these tools work together seamlessly. This vignette focuses on the data wrangling side: reading data, reshaping it into tidy form, and manipulating it with `dplyr`.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

---

## Tidy data

**Tidy data** is a standard way of organizing a dataset so that it is easy to analyze and visualize. The rules are simple:

1. Each **variable** has its own column
2. Each **observation** has its own row
3. Each **value** has its own cell

This might seem obvious, but most data you encounter in the wild is *not* tidy. Recognizing untidy data — and knowing how to fix it — is a core skill.

### Example: a gene expression timecourse

Here is a typical untidy representation of gene expression data. Each timepoint is a separate column, but `t0`, `t1`, `t2` are not variables — they are *values* of a `timepoint` variable.

```{r}
gene_expression <- tibble(
    Gene      = c("ACT1", "GAP1", "MEP2", "DUR3", "MSN2", "DAL80"),
    t0        = c(2.2,  4.3,  1.6, -1.2, -2.0,  0.2),
    t1        = c(3.2,  2.1,  0.8, -1.8, -0.8,  0.6),
    t2        = c(4.5,  1.6,  0.4, -1.8, -0.1,  0.9)
)
gene_expression
```

The tidy version has one row per gene per timepoint:

```{r}
gene_expression_tidy <- gene_expression |>
    pivot_longer(cols = t0:t2, names_to = "timepoint", values_to = "log2_expression")
gene_expression_tidy
```

This format is what `ggplot2` and `dplyr` expect. For example, plotting all genes across time now requires no data reshaping:

```{r}
ggplot(gene_expression_tidy, aes(x = timepoint, y = log2_expression,
                                  color = Gene, group = Gene)) +
    geom_point() +
    geom_line() +
    theme_classic() +
    labs(x = "Timepoint", y = "log2 expression", color = "Gene")
```

### Example: flow cytometry data

Here is a tidy flow cytometry dataset — each row is one measurement of one sample.

```{r}
facs_data <- tibble(
    Sample  = c("A1", "A1", "A2", "A2", "A3", "A3"),
    Channel = c("FSC", "FL1", "FSC", "FL1", "FSC", "FL1"),
    Value   = c(3.6, 4.5, 3.5, 3.2, 3.8, 4.2)
)
facs_data
```

But if we want to *correlate* FSC and FL1, we need them as separate columns — one row per sample. `pivot_wider()` does this:

```{r}
facs_data |> pivot_wider(names_from = Channel, values_from = Value)
```

### The built-in `table1`, `table2`, `table3`

`tidyverse` ships with example datasets illustrating tidy and untidy forms of the same data:

```{r}
table1   # tidy: one row per country per year
table2   # untidy: cases and population are values in a "type" column, not separate columns
table3   # untidy: rate is cases/population stored as a string in one column
```

To convert `table2` to tidy format:

```{r}
table2 |> pivot_wider(names_from = type, values_from = count)
```

---

## The pipe: `|>` and `%>%`

The pipe passes the result of one function directly into the next. It makes code read as a left-to-right sequence of steps, rather than nested function calls.

R 4.1 introduced the **native pipe** `|>` as a built-in language feature. You may also see `%>%` from the `magrittr` package (included in the tidyverse), which works similarly. Prefer `|>` for new code — it requires no package and is slightly faster.

```{r}
# Without the pipe — reads inside-out
summarise(filter(gene_expression_tidy, Gene == "GAP1"), mean_expr = mean(log2_expression))

# With the pipe — reads left to right
gene_expression_tidy |>
    filter(Gene == "GAP1") |>
    summarise(mean_expr = mean(log2_expression))
```

The pipe is pronounced **"then"**: *take gene_expression_tidy, then filter for GAP1, then summarize*.

---

## Reading data

`readr` functions read files into tibbles. The most common:

| Function | Use |
|---|---|
| `read_csv()` | Comma-delimited |
| `read_tsv()` | Tab-delimited |
| `read_delim()` | Any delimiter (specify with `delim =`) |

Read the *S. cerevisiae* R64 genome annotation (GFF3 format). We skip the 24-line comment header and assign column names manually, since GFF3 files have no header row.

```{r message=FALSE, warning=FALSE}
gff <- read_delim("../data/Saccharomyces_cerevisiae.R64-1-1.34.gff3",
    delim = "\t", col_names = FALSE,
    comment = "#", trim_ws = TRUE, skip = 24)

names(gff) <- c("chromosome", "source", "feature",
                "start", "stop", "score", "strand", "phase", "info")
```

### Inspecting a tibble

`glimpse()` is the tidyverse equivalent of `str()` — it shows column names, types, and the first few values in a compact format.

```{r}
glimpse(gff)
```

```{r}
# Number of rows and columns
nrow(gff)
ncol(gff)
```

Set categorical columns to factors, which is important for `dplyr` grouping and `ggplot2` plotting:

```{r}
gff <- gff |>
    mutate(
        feature    = as.factor(feature),
        chromosome = as.factor(chromosome),
        strand     = as.factor(strand)
    )
glimpse(gff)
```

---

## Core dplyr verbs

`dplyr` provides a small set of functions ("verbs") that cover most data manipulation tasks. They all take a tibble as the first argument and return a tibble, which makes them composable with the pipe.

### `select()` — choose columns

```{r}
gff |> select(chromosome, feature, start, stop, strand)
```

Drop columns with `-`:

```{r}
gff |> select(-source, -score, -phase, -info)
```

### `filter()` — subset rows

```{r}
# Keep only CDS features
gff |> filter(feature == "CDS")
```

```{r}
# Multiple conditions with & (AND) or | (OR)
gff |> filter(feature == "CDS" & chromosome == "I")
```

```{r}
# %in% is cleaner than chained | for membership tests
gff |> filter(feature %in% c("CDS", "tRNA_gene", "snoRNA"))
```

### `mutate()` — add or transform columns

```{r}
gff <- gff |> mutate(length = abs(stop - start))
gff |> select(chromosome, feature, length)
```

`mutate()` can create multiple columns in one call, and later expressions can reference columns created earlier in the same call:

```{r}
gff |>
    mutate(
        length    = abs(stop - start),
        length_kb = length / 1000
    ) |>
    select(feature, length, length_kb)
```

### `arrange()` — sort rows

```{r}
# Smallest features first
gff |> arrange(length) |> select(feature, chromosome, length)
```

```{r}
# Largest first with desc()
gff |> arrange(desc(length)) |> select(feature, chromosome, length)
```

### `summarise()` — compute summary statistics

`summarise()` collapses the tibble to one row of summary statistics:

```{r}
gff |>
    summarise(
        mean_length = mean(length),
        sd_length   = sd(length),
        min_length  = min(length),
        max_length  = max(length),
        n           = n()
    )
```

### `group_by()` + `summarise()` — summary by group

Combining `group_by()` with `summarise()` is one of the most common patterns in data analysis. It computes summary statistics separately for each level of the grouping variable.

```{r}
gff |>
    group_by(feature) |>
    summarise(
        mean_length = mean(length),
        sd_length   = sd(length),
        n           = n()
    ) |>
    arrange(desc(mean_length))
```

Always `ungroup()` after grouped operations if you're going to pass the result downstream — otherwise later operations may behave unexpectedly.

```{r}
gff |>
    filter(feature %in% c("CDS", "tRNA_gene", "snoRNA", "snRNA", "rRNA")) |>
    group_by(feature, chromosome) |>
    summarise(n = n(), mean_length = mean(length), .groups = "drop") |>
    arrange(chromosome, feature)
```

### Piping into ggplot2

The pipe works seamlessly with `ggplot2`. Just replace the final `|>` with `+` when switching to ggplot layers.

```{r message=FALSE, warning=FALSE}
gff |>
    filter(feature %in% c("CDS", "tRNA_gene", "snoRNA", "snRNA", "rRNA")) |>
    mutate(feature = fct_reorder(feature, length, .fun = median)) |>
    ggplot(aes(x = feature, y = length, fill = feature)) +
    geom_violin(show.legend = FALSE) +
    scale_y_log10() +
    scale_fill_viridis_d() +
    labs(x = "Feature type", y = "Length (bp)",
         title = "S. cerevisiae feature length distributions") +
    theme_classic()
```

---

## Reshaping data: `pivot_longer()` and `pivot_wider()`

`pivot_longer()` and `pivot_wider()` are the modern replacements for the deprecated `gather()` and `spread()`.

### `pivot_longer()` — wide to long

Use when column names encode a variable (e.g., each timepoint is a column). Makes the table longer and is the direction most commonly needed to get data into tidy format.

```{r}
gene_expression            # wide — columns are timepoints
```

```{r}
gene_expression |>
    pivot_longer(cols = t0:t2, names_to = "timepoint", values_to = "log2_expression")
```

`cols` selects which columns to pivot. You can use any `select()`-style selector: `t0:t2`, `starts_with("t")`, `where(is.numeric)`, etc.

### `pivot_wider()` — long to wide

Use when one observation is spread across multiple rows. Makes the table wider.

```{r}
facs_data                  # long — FSC and FL1 are in rows
```

```{r}
facs_data |> pivot_wider(names_from = Channel, values_from = Value)
```

### Correlation plots require wide format

Tidy (long) format is great for `ggplot2`, but scatter plots comparing two variables require them to be in separate columns. Go wide, plot, then go back to long if needed.

```{r}
gene_expression_tidy |>
    pivot_wider(names_from = timepoint, values_from = log2_expression) |>
    ggplot(aes(x = t0, y = t1, label = Gene)) +
    geom_point(size = 3, color = "steelblue") +
    ggrepel::geom_text_repel() +  # install ggrepel if needed: install.packages("ggrepel")
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    theme_classic() +
    labs(x = "t0 expression", y = "t1 expression")
```

---

## String manipulation with `stringr`

`stringr` is part of the tidyverse and provides consistent, readable functions for working with character strings. All functions start with `str_`.

| Function | What it does |
|---|---|
| `str_detect(x, pattern)` | Returns TRUE/FALSE if pattern is found |
| `str_extract(x, pattern)` | Extracts the first match |
| `str_replace(x, pattern, replacement)` | Replaces first match |
| `str_replace_all(x, pattern, replacement)` | Replaces all matches |
| `str_remove(x, pattern)` | Removes first match |
| `str_split(x, pattern)` | Splits into a list |
| `str_to_lower()` / `str_to_upper()` | Case conversion |

Patterns follow **regex** syntax. A few useful patterns:
- `"^text"` — starts with "text"
- `"text$"` — ends with "text"
- `"(?<=Name=)[^;]+"` — everything after "Name=" up to the next ";"

### Extracting gene names from the GFF3 info column

The `info` column in GFF3 files is a semicolon-separated list of key=value pairs. We can use regex with `str_extract()` to pull out specific fields cleanly.

```{r}
gff |> filter(feature == "gene") |> select(info) |> head()
```

```{r warning=FALSE}
gff_genes <- gff |>
    filter(feature == "gene") |>
    mutate(
        systematic_name = str_extract(info, "(?<=ID=)[^;]+"),
        gene_name       = str_extract(info, "(?<=Name=)[^;]+"),
        description     = str_extract(info, "(?<=description=)[^;]+"),
        # Decode URL-encoded characters common in GFF3 descriptions
        description     = str_replace_all(description, "%3B", ";"),
        description     = str_replace_all(description, "%2C", ","),
        description     = str_replace_all(description, "%20", " ")
    ) |>
    select(chromosome, systematic_name, gene_name, start, stop, length, strand, description)

gff_genes
```

### Writing results to a file

Use `write_csv()` or `write_tsv()` to save tibbles. Prefer these over base R `write.csv()` — they don't add row numbers and handle NA and special characters consistently.

```{r eval=FALSE}
write_tsv(gff_genes, "yeast_genes_annotated.tsv")
```

---

## Joining tables

Real analyses almost always involve combining data from multiple sources. `dplyr` join functions match rows between two tables on a shared key column.

### Types of joins

| Function | Result |
|---|---|
| `left_join(x, y)` | All rows from x; matching rows from y (most common) |
| `inner_join(x, y)` | Only rows with a match in both x and y |
| `full_join(x, y)` | All rows from both x and y |
| `anti_join(x, y)` | Rows in x that have **no** match in y (useful for finding unmatched entries) |

### Example: joining genome annotations with expression data

Suppose we have RNA-seq differential expression results and want to annotate them with gene descriptions from our `gff_genes` table.

```{r}
# Simulate some differential expression results
set.seed(42)
deg_results <- tibble(
    systematic_name = sample(gff_genes$systematic_name, 20),
    log2FC          = rnorm(20, mean = 0, sd = 1.5),
    padj            = runif(20, min = 0, max = 0.1)
)
deg_results
```

```{r}
# Left join: keep all DEG results, add annotation columns from gff_genes
deg_annotated <- deg_results |>
    left_join(gff_genes |> select(systematic_name, gene_name, description, chromosome),
              by = "systematic_name")
deg_annotated
```

`anti_join()` is useful for quality control — finding rows in one table that failed to match:

```{r}
# Any DEG results that didn't match a gene annotation?
deg_results |>
    anti_join(gff_genes, by = "systematic_name")
```

---

## Best practices

**Use `|>` instead of `%>%`** for new code. The native pipe requires no package, is slightly faster, and is the direction the language is moving.

**Name your intermediate objects clearly.** Break long pipelines into named steps (`gff_filtered`, `gff_genes`) rather than building one enormous pipe. Easier to debug and re-use.

**Use `.groups = "drop"` in `summarise()`** after `group_by()` to avoid leaving the result in a grouped state, which can cause surprising behavior downstream.

**Prefer `pivot_longer()` / `pivot_wider()` over `gather()` / `spread()`.** The latter are deprecated. The new functions have clearer argument names and handle more cases.

**Use `str_extract()` with regex** rather than chains of `separate()` for parsing structured strings like GFF3 info fields. More concise and less fragile.

**Check joins with `anti_join()`.** Before trusting a join result, verify that the key columns actually match using `anti_join()`. Missing matches are a common source of silent bugs.

**Use `fct_reorder()` before plotting factors.** Sorting categorical variables by a summary statistic (median, count) makes plots much easier to read.

---

## Exercises

### Exercise 1
Using `gff_genes`, find the 10 longest genes on chromosome IV. Display only the `gene_name`, `systematic_name`, and `length` columns.

```{r}
# Your code here
gff_genes |> filter(chromosome == "IV")
```

### Exercise 2
Using `gff`, calculate the number of features of each type (`feature`) on each chromosome. Then filter to keep only chromosomes I–IV and plot the result as a stacked bar chart colored by feature type.

```{r message=FALSE, warning=FALSE}
# Your code here
gff |> group_by(chromosome, feature) |> summarise(n = n(), .groups = "drop")
```

### Exercise 3
The `gene_expression` tibble has expression values at t0, t1, and t2. Pivot it to long format, then for each gene calculate the change in expression from t0 to t2 (i.e., t2 − t0). Which gene changed the most?

```{r}
# Your code here
gene_expression
```

### Exercise 4
Join `gff_genes` with `deg_annotated` to get the chromosomal location of each differentially expressed gene. Then make a bar plot showing how many DEGs are on each chromosome.

```{r}
# Your code here
deg_annotated
```

---

## Resources

- [R for Data Science (2nd ed.)](https://r4ds.hadley.nz/) — the definitive tidyverse textbook, freely available online
- [dplyr cheatsheet](https://rstudio.github.io/cheatsheets/html/data-transformation.html)
- [tidyr cheatsheet](https://rstudio.github.io/cheatsheets/html/tidyr.html)
- [stringr cheatsheet](https://rstudio.github.io/cheatsheets/html/strings.html)
- [Tidy data paper](https://www.jstatsoft.org/article/view/v059i10) (Wickham, 2014) — the original paper defining tidy data principles
