---
title: "Building a custom OrgDB and running GO overrepresentation analysis"
author: "Ozan Berk Imir & Ernesto Segredo Otero"
date: "Last compiled on `r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Introduction

An **OrgDB** (Organism Annotation Database) is a Bioconductor object that stores genome-wide mappings between gene identifiers, gene symbols, GO terms, chromosomal locations, and other annotations. The Bioconductor project provides pre-built OrgDBs for common model organisms (e.g., `org.Sc.sgd.db` for *S. cerevisiae*, `org.Hs.eg.db` for human), but for less-studied organisms you must build one yourself.

This vignette demonstrates how to:

1. Build a custom OrgDB for *Candida albicans* using `AnnotationForge`, by merging data from UniProt and the Candida Genome Database (CGD)
2. Build an OrgDB for organisms where UniProt already provides GO annotations (illustrated with *Mustela putorius*)
3. Run a GO overrepresentation analysis on DESeq2 results using the custom OrgDB and `clusterProfiler`

## Required packages

All packages are on Bioconductor. Install once if needed:

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("AnnotationForge", "clusterProfiler", "enrichplot"))
install.packages("wordcloud")
```

```{r libraries}
library(tidyverse)
library(AnnotationForge)
library(clusterProfiler)
library(enrichplot)
library(wordcloud)
```

---

## Part 1: Building an OrgDB for *Candida albicans*

### Data sources

`makeOrgPackage()` requires three data frames with specific column structures. We build these by merging two external data sources:

- **UniProt** — protein and gene information for all *C. albicans* strains. Download from [UniProt](https://www.uniprot.org/) by searching for *Candida albicans*, then using "Customize columns" to select: Entry, Gene names, Gene names (primary), Gene names (ordered locus), Protein names, Reviewed, Entry name, Organism, Length. Export as TSV.
- **GO data** — GO term associations from the [Candida Genome Database](http://www.candidagenome.org/download/go/). Download the GAF-format file (`gene_association.cgd`).

### Reading and merging the data

```{r read-data}
# Read the UniProt TSV export (header row present)
uniprot <- read_tsv("../data/uniprotkb_candida_albicans_2024_07_16.tsv",
                    comment = "!")

# Read the CGD GO annotation file (GAF 2.0 format; no header, '!' comment lines)
GO_data <- read_tsv("../data/gene_association.cgd",
                    col_names = FALSE, comment = "!")
```

The two files share gene symbols (e.g., AAF1) — `Gene Names (primary)` in UniProt and `V3` in the GO file. We join on these, then extract the 9-character CGD locus ID (e.g., `C3_06470W`) from column `V11` of the GO file, which lists gene synonyms as a pipe-separated string. IDs that do not match the *C. albicans* SC5314 locus format (two characters + underscore) belong to other *Candida* species in the file and are removed.

```{r merge-data}
# Inner join: keep only genes present in both files
merged_data <- inner_join(uniprot, GO_data,
                          by = c("Gene Names (primary)" = "V3"))

# Extract the 9-character locus ID from the synonyms column (V11)
# V11 is a pipe-separated list; the C. albicans locus ID is the entry of the form C#_#####N
merged_data <- merged_data |>
    mutate(
        gene = substr(V11, 1, 9),
        # Locus IDs for C. albicans SC5314 have an underscore at position 3 (e.g., C3_06470W)
        # IDs from other Candida species do not; set those to NA and remove
        gene = if_else(grepl("^.{2}[^_]", gene), NA_character_, gene)
    ) |>
    filter(!is.na(gene))
```

### Preparing the three required data frames

`makeOrgPackage()` expects three data frames, each with a `GID` column as the primary key:

- **candidaSym** — `GID`, `SYMBOL`, `GENENAME`
- **candidaChr** — `GID`, `CHROMOSOME`
- **candidaGO**  — `GID`, `GO`, `EVIDENCE`

#### SYMBOL data frame

`SYMBOL` is the short gene name (e.g., AAF1); `GENENAME` is the full protein name from UniProt. Long protein names are truncated to the first five words with any parenthetical text removed.

```{r sym-df}
candidaSym <- merged_data |>
    select(GID      = gene,
           SYMBOL   = `Gene Names (primary)`,
           GENENAME = `Protein names`) |>
    filter(!is.na(SYMBOL), !is.na(GENENAME)) |>
    distinct(SYMBOL,   .keep_all = TRUE) |>
    distinct(GENENAME, .keep_all = TRUE) |>
    mutate(
        GENENAME = sapply(strsplit(GENENAME, " "), function(x) {
            first_five <- paste(x[seq_len(min(5, length(x)))], collapse = " ")
            gsub("\\(.*", "", first_five)
        })
    )

head(candidaSym)
```

#### CHROMOSOME data frame

*C. albicans* has 8 chromosomes (7 autosomal + R). The chromosome number is the second character of the locus ID (e.g., `C3_06470W` → chromosome 3).

```{r chr-df}
candidaChr <- merged_data |>
    select(GID = gene) |>
    mutate(
        CHROMOSOME = substr(GID, 2, 2),
        CHROMOSOME = if_else(grepl("[0-9R]", CHROMOSOME), CHROMOSOME, NA_character_)
    ) |>
    filter(!is.na(CHROMOSOME)) |>
    distinct()

head(candidaChr)
```

#### GO data frame

In GAF format, column `V5` is the GO term and `V7` is the evidence code.

```{r go-df}
candidaGO <- merged_data |>
    select(GID      = gene,
           GO       = V5,
           EVIDENCE = V7) |>
    filter(!is.na(GO), !is.na(EVIDENCE)) |>
    distinct()

head(candidaGO)
```

### Creating and installing the OrgDB package

> **Important:** `makeOrgPackage()` creates a new package directory in your working directory. It will throw an error if that directory already exists. **Run the `makeOrgPackage()` call only once**, then comment it out. The `install.packages()` call below it should be run every time on a new machine.

```{r make-orgdb, eval=FALSE}
# Run ONCE — comment out makeOrgPackage() after the first run
makeOrgPackage(
    gene_info  = candidaSym,
    chromosome = candidaChr,
    go         = candidaGO,
    version    = "0.1",
    maintainer = "Ozan Imir <obi203@nyu.edu>",
    author     = "Ozan Imir <obi203@nyu.edu>",
    outputDir  = ".",
    tax_id     = "237561",
    genus      = "Candida",
    species    = "albicans",
    goTable    = "go"
)

# Install the generated package from source
install.packages("./org.Calbicans.eg.db", type = "source", repos = NULL)
```

```{r load-orgdb}
library(org.Calbicans.eg.db)

# Verify available key types
columns(org.Calbicans.eg.db)
```

---

## Alternative: OrgDB when UniProt provides GO annotations

For organisms where UniProt already includes GO term associations, you can skip the separate GO database file. You still need chromosome information (from NCBI) and GO evidence codes (from QuickGO). This example uses *Mustela putorius* (ferret, taxonomy ID 9669).

> **Note:** The data files for this section are organism-specific and not included in the repository. Download them as described below before running these chunks.

**Data sources:**
- **UniProt** — search by taxonomy ID 9669; export Entry, Gene Names, Protein names, Gene Ontology IDs as TSV
- **NCBI** — download gene table including chromosome information
- **QuickGO** — download GO evidence codes filtered by organism

```{r weasel-read, eval=FALSE}
mi_uniprot <- read_tsv("uniprotkb_taxonomy_id_9669_2024_10_01.tsv", comment = "!")
mi_ncbi    <- read_tsv("mustela_putorius_furo_ncbi.tsv",             comment = "!")
mi_quickGO <- read_tsv("QuickGO-annotations.tsv",                    comment = "!") |>
    mutate(GeneGO = paste0(`GENE PRODUCT ID`, "_", `GO TERM`))
```

```{r weasel-sym, eval=FALSE}
miSym <- mi_uniprot |>
    select(GID = Entry, SYMBOL = `Gene Names`, GENENAME = `Protein names`) |>
    filter(SYMBOL != "", GENENAME != "") |>
    distinct()
```

```{r weasel-chr, eval=FALSE}
miChr <- mi_ncbi |>
    select(GID = Symbol, CHROMOSOME = Chromosome) |>
    filter(GID != "") |>
    mutate(CHROMOSOME = if_else(CHROMOSOME == "", "UNKNOWN", CHROMOSOME)) |>
    distinct()
```

```{r weasel-go, eval=FALSE}
# UniProt stores multiple GO IDs per row, separated by "; " — pivot to one row per gene-GO pair
miGO_long <- mi_uniprot |>
    select(GID = Entry, GO = `Gene Ontology IDs`) |>
    filter(GO != "") |>
    distinct() |>
    separate_rows(GO, sep = ";\\s*") |>
    mutate(GeneGO = paste0(GID, "_", GO)) |>
    left_join(mi_quickGO |> select(GeneGO, EVIDENCE = `GO EVIDENCE CODE`),
              by = "GeneGO") |>
    mutate(EVIDENCE = replace_na(EVIDENCE, "UNKNOWN")) |>
    select(GID, GO, EVIDENCE) |>
    distinct()
```

```{r weasel-orgdb, eval=FALSE}
# Run ONCE — comment out makeOrgPackage() after the first run
makeOrgPackage(
    gene_info  = miSym,
    chromosome = miChr,
    go         = miGO_long,
    version    = "1.0",
    maintainer = "Ernesto Segredo Otero <es6393@nyu.edu>",
    author     = "Ernesto Segredo Otero <es6393@nyu.edu>",
    outputDir  = ".",
    tax_id     = "9669",
    genus      = "Mustela",
    species    = "putoriusNew",
    goTable    = "go"
)

install.packages("./org.MputoriusNew.eg.db", repos = NULL, type = "source")
library(org.MputoriusNew.eg.db)

columns(org.MputoriusNew.eg.db)
keytypes(org.MputoriusNew.eg.db)
```

---

## Part 2: GO overrepresentation analysis with clusterProfiler

This section uses the *C. albicans* OrgDB built in Part 1 to run a GO overrepresentation analysis on DESeq2 differential expression results.

### Reading DESeq2 results

```{r read-deseq}
# DESeq2 output — columns: Locus, log2FoldChange, padj
df <- read_csv("../data/time_72_vs_2_results.csv")
glimpse(df)
```

### Preparing the gene lists

`enrichGO()` requires two inputs:

- **`gene`** — the genes of interest (significant DEGs)
- **`universe`** — all tested genes (the background set)

```{r prep-genes}
# Universe: all tested genes, named vector of log2FC sorted in decreasing order
gene_list <- df |>
    filter(!is.na(log2FoldChange)) |>
    arrange(desc(log2FoldChange)) |>
    pull(log2FoldChange, name = Locus)

# Gene set of interest: significant DEGs with |log2FC| > 2
genes <- df |>
    filter(padj < 0.05, abs(log2FoldChange) > 2, !is.na(log2FoldChange)) |>
    pull(Locus)

cat("Universe size:", length(gene_list), "\n")
cat("Gene set size:", length(genes), "\n")
```

### Creating the enrichGO object

The `ont` parameter selects the GO sub-ontology: `"BP"` (Biological Process), `"MF"` (Molecular Function), `"CC"` (Cellular Component), or `"ALL"`. The `keyType` must match an available key type in your OrgDB — check with `keytypes(org.Calbicans.eg.db)`.

```{r enrich-go}
go_enrich <- enrichGO(
    gene         = genes,
    universe     = names(gene_list),
    OrgDb        = org.Calbicans.eg.db,
    keyType      = "GID",
    readable     = TRUE,
    ont          = "ALL",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2
)
```

### Visualizing results

#### Results table

```{r results-table}
head(go_enrich)
```

#### Barplot — top enriched terms by gene count

```{r barplot, fig.width=8, fig.height=5}
barplot(go_enrich, showCategory = 10, title = "GO Enrichment", font.size = 8)
```

#### Dotplot — enrichment score vs gene ratio

```{r dotplot, fig.width=8, fig.height=6}
dotplot(go_enrich)
```

#### Upset plot — gene overlap across enriched terms

```{r upsetplot, fig.width=18, fig.height=12}
upsetplot(go_enrich)
```

#### Enrichment map — network of overlapping gene sets

Enriched terms are connected when their gene sets overlap. Functionally related terms cluster together, making it easy to identify broader biological themes.

```{r emapplot, fig.width=10, fig.height=8}
pairwise_go_enrich <- pairwise_termsim(go_enrich)
emapplot(pairwise_go_enrich,
         layout.params = list(layout = "star"),
         showCategory  = 25,
         cex.params    = list(category_label = 0.6),
         shadowtext    = TRUE)
```

#### Category netplot — genes linked to enriched terms

Shows which individual genes drive each enriched term, and highlights genes belonging to multiple categories.

```{r cnetplot, fig.width=12, fig.height=8}
cnetplot(go_enrich,
         categorySize = "pvalue",
         color.params = list(foldChange = gene_list))
```

#### Word cloud — term frequency by gene ratio

```{r wordcloud, fig.width=14, fig.height=12}
wc_df <- data.frame(
    term = go_enrich$Description,
    freq = as.integer(gsub("/.*", "", go_enrich$GeneRatio))
)
wordcloud(words     = wc_df$term,
          freq      = wc_df$freq,
          scale     = c(4, 0.1),
          colors    = brewer.pal(8, "Dark2"),
          max.words = 25)
```

---

## Session info

```{r session-info}
sessioninfo::session_info()
```
