---
title: "Building a custom OrgDB and running GO overrepresentation analysis"
author: "Ozan Berk Imir & Ernesto Segredo Otero"
date: "Last compiled on `r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)  # all chunks require external data; set eval=TRUE to run
```

## Introduction

An **OrgDB** (Organism Annotation Database) is a Bioconductor object that stores genome-wide mappings between gene identifiers, gene symbols, GO terms, chromosomal locations, and other annotations. The Bioconductor project provides pre-built OrgDBs for common model organisms (e.g., `org.Sc.sgd.db` for *S. cerevisiae*, `org.Hs.eg.db` for human), but for less-studied organisms you must build one yourself.

This vignette demonstrates how to:

1. Build a custom OrgDB for *Candida albicans* using `AnnotationForge`, by merging data from UniProt and the Candida Genome Database (CGD)
2. Run a GO overrepresentation analysis on DESeq2 results using the custom OrgDB and `clusterProfiler`
3. Build an OrgDB for organisms where UniProt already provides GO annotations (illustrated with *Mustela putorius*)

## Required packages

All packages are on Bioconductor. Install once if needed:

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("AnnotationForge", "clusterProfiler", "enrichplot"))
install.packages("wordcloud")
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(AnnotationForge)
library(clusterProfiler)
library(enrichplot)
library(wordcloud)
```

---

## Part 1: Building an OrgDB for *Candida albicans*

### Data sources

`makeOrgPackage()` requires three data frames with specific column structures. We build these by merging two external data sources:

- **UniProt** — protein and gene information for all *C. albicans* strains. Download from [UniProt](https://www.uniprot.org/) by searching for *Candida albicans*, then using "Customize columns" to select: Entry, Gene names, Gene names (primary), Gene names (ordered locus), Protein names, Reviewed, Entry name, Organism, Length. Export as TSV.
- **GO data** — GO term associations from the [Candida Genome Database](http://www.candidagenome.org/download/go/). Download the GAF-format file (`gene_association.cgd`).

### Reading and merging the data

```{r read-data}
# Read the UniProt TSV export
uniprot <- read_tsv("uniprotkb_candida_albicans_2024_07_16.tsv",
                    comment = "!")

# Read the CGD GO annotation file (GAF format; no header, '!' comment lines)
GO_data <- read_tsv("gene_association.cgd",
                    col_names = FALSE, comment = "!")
```

The two files share gene symbols (e.g., GAP1) — `Gene.Names..primary.` in UniProt and `V3` in the GO file. We merge on these, then extract the 9-character CGD locus ID (e.g., `C1_00010W`) from column V11 of the GO file. IDs that don't match the expected format belong to other *Candida* strains and are removed.

```{r merge-data}
# Merge on shared gene symbol
merged_data <- merge(uniprot, GO_data,
                     by.x = "Gene.Names..primary.",
                     by.y = "V3")

# Extract the 9-character locus ID from the GO data gene ID column (V11)
# CGD locus IDs follow the pattern: two letters, underscore, five digits, letter (e.g. C1_00010W)
merged_data <- merged_data |>
    mutate(
        gene = substr(V11, 1, 9),
        # IDs that don't start with two chars + underscore belong to other strains — set to NA
        gene = if_else(grepl("^.{2}[^_]", gene), NA_character_, gene)
    ) |>
    filter(!is.na(gene))
```

### Preparing the three required data frames

`makeOrgPackage()` expects three data frames, each containing a `GID` column as the primary key:

- **candidaSym** — `GID`, `SYMBOL`, `GENENAME`
- **candidaChr** — `GID`, `CHROMOSOME`
- **candidaGO**  — `GID`, `GO`, `EVIDENCE`

#### SYMBOL data frame

```{r sym-df}
# Select GID, primary gene symbol, and protein name columns by name
# Adjust column names if your UniProt export differs
candidaSym <- merged_data |>
    select(GID = gene,
           SYMBOL = `Gene.Names..primary.`,
           GENENAME = `Protein.names`) |>
    filter(!is.na(SYMBOL), !is.na(GENENAME)) |>
    distinct(SYMBOL, .keep_all = TRUE) |>
    distinct(GENENAME, .keep_all = TRUE) |>
    mutate(
        # Truncate long protein names: keep first 5 words, drop parenthetical text
        GENENAME = sapply(strsplit(GENENAME, " "), function(x) {
            first_five <- paste(x[seq_len(min(5, length(x)))], collapse = " ")
            gsub("\\(.*", "", first_five)
        })
    )
```

#### CHROMOSOME data frame

*C. albicans* has 8 chromosomes (7 autosomal + R chromosome). The chromosome number is encoded in the locus ID (second character).

```{r chr-df}
# Select GID and chromosome info; extract chromosome number from locus ID
candidaChr <- merged_data |>
    select(GID = gene, CHROMOSOME = V19) |>   # adjust V19 if column differs
    filter(!is.na(CHROMOSOME)) |>
    mutate(
        # Extract chromosome: second character of locus ID
        CHROMOSOME = substr(CHROMOSOME, 2, 2),
        # Keep only valid chromosome labels (digits and R); set others to NA
        CHROMOSOME = if_else(grepl("[0-9R]", CHROMOSOME), CHROMOSOME, NA_character_)
    ) |>
    filter(!is.na(CHROMOSOME)) |>
    distinct()
```

#### GO data frame

```{r go-df}
# Select GID, GO term, and evidence code columns
candidaGO <- merged_data |>
    select(GID = gene,
           GO = V5,          # GO term column in GAF format
           EVIDENCE = V7) |> # evidence code column in GAF format
    filter(!is.na(GO), !is.na(EVIDENCE)) |>
    distinct()
```

### Creating and installing the OrgDB package

> **Important:** `makeOrgPackage()` creates a new directory in your working directory. It will error if that directory already exists, so **run this chunk only once**. After running it, set `eval=FALSE` on this chunk or comment out the `makeOrgPackage()` call.

```{r make-orgdb}
# Run ONCE — comment out makeOrgPackage() after the first run
makeOrgPackage(
    gene_info  = candidaSym,
    chromosome = candidaChr,
    go         = candidaGO,
    version    = "0.1",
    maintainer = "Ozan Imir <obi203@nyu.edu>",
    author     = "Ozan Imir <obi203@nyu.edu>",
    outputDir  = ".",
    tax_id     = "237561",
    genus      = "Candida",
    species    = "albicans",
    goTable    = "go"
)

# Install the generated package from source — run every time on a new machine
install.packages("./org.Calbicans.eg.db", type = "source", repos = NULL)
```

```{r load-orgdb}
library(org.Calbicans.eg.db)

# Verify the package loaded correctly
columns(org.Calbicans.eg.db)
keytypes(org.Calbicans.eg.db)
```

---

## Part 2: GO overrepresentation analysis with clusterProfiler

This section uses the *C. albicans* OrgDB built above to run a GO overrepresentation analysis on DESeq2 differential expression results.

### Reading DESeq2 results

```{r read-deseq}
# Read DESeq2 output — expects columns: Locus, log2FoldChange, padj
df <- read_csv("data/time_72_vs_2_results.csv")
```

### Preparing the gene lists

`enrichGO()` requires two inputs:
- **`gene`** — the genes of interest (e.g., significant DEGs)
- **`universe`** — all tested genes (the background set)

```{r prep-genes}
# Universe: all tested genes, named by locus ID, sorted by log2FC (required by clusterProfiler)
gene_list <- df |>
    filter(!is.na(log2FoldChange)) |>
    arrange(desc(log2FoldChange)) |>
    pull(log2FoldChange, name = Locus)

# Gene set of interest: significant DEGs with |log2FC| > 2
genes <- df |>
    filter(padj < 0.05, abs(log2FoldChange) > 2, !is.na(log2FoldChange)) |>
    pull(Locus)
```

### Creating the enrichGO object

The `ont` parameter selects the GO sub-ontology:

- `"BP"` — Biological Process
- `"MF"` — Molecular Function
- `"CC"` — Cellular Component
- `"ALL"` — all three (results include an `ONTOLOGY` column)

The `keyType` must match an available key type in your OrgDB. Check with `keytypes(org.Calbicans.eg.db)`.

```{r enrich-go}
go_enrich <- enrichGO(
    gene          = genes,
    universe      = names(gene_list),
    OrgDb         = org.Calbicans.eg.db,
    keyType       = "GID",
    readable      = TRUE,
    ont           = "ALL",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.2
)
```

### Visualizing results

#### Results table

```{r results-table}
head(go_enrich)
```

#### Barplot — top enriched terms by gene count

```{r barplot, fig.width=8, fig.height=5}
barplot(go_enrich,
        showCategory = 10,
        title = "GO Enrichment",
        font.size = 8)
```

#### Dotplot — enrichment score vs gene ratio

```{r dotplot, fig.width=8, fig.height=6}
dotplot(go_enrich)
```

#### Upset plot — gene overlap across enriched terms

```{r upsetplot, fig.width=18, fig.height=12}
upsetplot(go_enrich)
```

#### Enrichment map — network of overlapping gene sets

Enriched terms are connected when their associated gene sets overlap. Functionally related terms cluster together, making it easy to identify broader biological themes.

```{r emapplot, fig.width=10, fig.height=8}
pairwise_go_enrich <- pairwise_termsim(go_enrich)
emapplot(pairwise_go_enrich,
         layout.params  = list(layout = "star"),
         showCategory   = 25,
         cex.params     = list(category_label = 0.6),
         shadowtext      = TRUE)
```

#### Category netplot — genes linked to enriched terms

The cnetplot shows which individual genes drive each enriched term, and highlights genes that belong to multiple categories.

```{r cnetplot, fig.width=12, fig.height=8}
cnetplot(go_enrich,
         categorySize = "pvalue",
         color.params = list(foldChange = gene_list))
```

#### Word cloud — term frequency by gene ratio

```{r wordcloud, fig.width=14, fig.height=12}
wc_df <- data.frame(
    term  = go_enrich$Description,
    freq  = as.integer(gsub("/.*", "", go_enrich$GeneRatio))
)
wordcloud(words  = wc_df$term,
          freq   = wc_df$freq,
          scale  = c(4, 0.1),
          colors = brewer.pal(8, "Dark2"),
          max.words = 25)
```

---

## Alternative: OrgDB when UniProt provides GO annotations

For organisms where UniProt already includes GO term associations, you can skip the CGD GO file. You still need chromosome information and GO evidence codes. In this example we use *Mustela putorius* (ferret), sourcing data from UniProt, NCBI, and QuickGO.

**Data sources:**
- **UniProt** — search by taxonomy ID 9669; export Entry, Gene Names, Protein names, Gene Ontology IDs
- **NCBI** — download gene table including chromosome information
- **QuickGO** — download GO evidence codes filtered by organism

```{r weasel-read}
mi_uniprot <- read_tsv("uniprotkb_taxonomy_id_9669_2024_10_01.tsv", comment = "!")
mi_ncbi    <- read_tsv("mustela_putorius_furo_ncbi.tsv",             comment = "!")
mi_quickGO <- read_tsv("QuickGO-annotations.tsv",                    comment = "!")

# Create a composite key for joining GO evidence codes
mi_quickGO <- mi_quickGO |>
    mutate(GeneGO = paste0(`GENE PRODUCT ID`, "_", `GO TERM`))
```

```{r weasel-sym}
# SYMBOL data frame
miSym <- mi_uniprot |>
    select(GID = Entry, SYMBOL = `Gene Names`, GENENAME = `Protein names`) |>
    filter(SYMBOL != "", GENENAME != "") |>
    distinct()
```

```{r weasel-chr}
# CHROMOSOME data frame
miChr <- mi_ncbi |>
    select(GID = Symbol, CHROMOSOME = Chromosome) |>
    filter(GID != "") |>
    mutate(CHROMOSOME = if_else(CHROMOSOME == "", "UNKNOWN", CHROMOSOME)) |>
    distinct()
```

```{r weasel-go}
# GO data frame: UniProt stores multiple GO IDs per row, separated by "; "
# pivot to one row per gene-GO pair, then join evidence codes from QuickGO
miGO_long <- mi_uniprot |>
    select(GID = Entry, GO = `Gene Ontology IDs`) |>
    filter(GO != "") |>
    distinct() |>
    separate_rows(GO, sep = ";\\s*") |>
    mutate(GeneGO = paste0(GID, "_", GO)) |>
    left_join(mi_quickGO |> select(GeneGO, EVIDENCE = `GO EVIDENCE CODE`),
              by = "GeneGO") |>
    mutate(EVIDENCE = replace_na(EVIDENCE, "UNKNOWN")) |>
    select(GID, GO, EVIDENCE) |>
    distinct()
```

```{r weasel-orgdb}
# Run ONCE — comment out makeOrgPackage() after the first run
makeOrgPackage(
    gene_info  = miSym,
    chromosome = miChr,
    go         = miGO_long,
    version    = "1.0",
    maintainer = "Ernesto Segredo Otero <es6393@nyu.edu>",
    author     = "Ernesto Segredo Otero <es6393@nyu.edu>",
    outputDir  = ".",
    tax_id     = "9669",
    genus      = "Mustela",
    species    = "putoriusNew",
    goTable    = "go"
)

install.packages("./org.MputoriusNew.eg.db", repos = NULL, type = "source")
```

```{r weasel-load}
library(org.MputoriusNew.eg.db)

columns(org.MputoriusNew.eg.db)
keytypes(org.MputoriusNew.eg.db)
```

---

## Session info

```{r session-info}
sessioninfo::session_info()
```
