---
title: "Analyze FACS data"
author: "David Gresham"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_notebook:
    toc: yes
    fig_caption: yes
    number_sections: yes
  html_document:
    toc: yes
    df_print: paged
---

This Vignette written as an [R Markdown](http://rmarkdown.rstudio.com) Notebook provides a simple example of importing FACS data from different machines, annotating the samples and visualizing the results.

The data are generated using the three FACS machines used in NYU CGSB gencore

1. Accuri
2. Aria
3. Cytek

# Example Data
Below are the data that are used in this vignette.  To work through this analysis:  

1. Download all directories and files from [here](https://drive.google.com/drive/folders/1_QARsH5vOHR-lIPURf3lgt6skN2of5US?usp=sharing)
2. Organize directories with the following structure.

You can see that there are three different directories `/Accuri` , `/Aria` and `/Cytek` that contain different .fcs files and samplesheets
```{bash}
echo '/Users/david/Projects/Data/flowdata'
ls -R /Users/david/Projects/Data/flowdata 
```

# Load required libraries

Several packages are required for running this analysis.  If they are not installed, you need to run the following commands the first time:

## Install Libraries
This workflow uses a variety of Bioconductor packages that should be loaded using: `BiocManager::install()` as per these [guidelines](https://www.bioconductor.org/install/).

```{r Install libraries, warning=FALSE}
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install()
#BiocManager::install(ncdfFlow)
#BiocManager::install(flowCore)
#BiocManager::install(flowViz)
#BiocManager::install(ggcyto)
#BiocManager::install(ggforce)


#install.packages(tidyverse)
#install.packages(ggridges)

```

## Loading the Gresham Lab's `flowtools` from [github](https://github.com/GreshamLab/flowtools)

```{r}
#install.packages(devtools)
#devtools::install_github("greshamlab/flowtools")
```

## Load libraries
This workflow using the following packages (include linkes to vignettes using packages)

* [flowCore](https://www.bioconductor.org/packages/release/bioc/vignettes/flowCore/inst/doc/HowTo-flowCore.pdf), which provides several functions for reading and manipulating .fcs files.
* [ncdfFlow](https://bioconductor.org/packages/devel/bioc/vignettes/ncdfFlow/inst/doc/ncdfFlow.pdf), which reads .fcs files to disk rather than memory.  This allows import of larger datasets as they do not need to be stored in memory.
* [ggcyto](https://www.bioconductor.org/packages/release/bioc/html/ggcyto.html)
    + [basic functions](https://www.bioconductor.org/packages/release/bioc/vignettes/ggcyto/inst/doc/Top_features_of_ggcyto.html)
    + [autoplot](https://www.bioconductor.org/packages/release/bioc/vignettes/ggcyto/inst/doc/autoplot.html)
    + [visualize flowset](https://www.bioconductor.org/packages/release/bioc/vignettes/ggcyto/inst/doc/ggcyto.flowSet.html)
    + [visualize gatingset](https://www.bioconductor.org/packages/release/bioc/vignettes/ggcyto/inst/doc/ggcyto.GatingSet.html)

```{r}
library(ncdfFlow)
library(flowCore)
library(flowViz)
library(ggcyto)
library(ggforce)
library(tidyverse)
library(ggridges)
library(readxl)
```

# Working with Accuri Data
First we will visualize flow data from the Accuri

## Read files
FCS files exported from the Accuri flow cytometry machine are read into R using `read.ncdfFlowSet`
```{r}
dir = '/Users/david/Projects/Data/flowdata/Accuri'
files <- sort(list.files(path=dir,pattern = ".fcs", full.names=TRUE))
flowData <- read.ncdfFlowSet(files=files, pattern=".fcs", alter.names = TRUE)

#needs to differ depending on sample sheet file typ
#sample.sheet <- read_excel(paste(path=dir,"GAP1_multicheck_20May2019.xlsx", sep="/"))
SampleSheet <- read_csv("~/Projects/Data/flowdata/Accuri/SampleSheet.csv")

#print(flowData)
#print(summary(flowData))
flowData[[1]]
```

## Assign sample names and other metadata
Information about the flowset is stored in an __AnnotatedDataFrame__.
The dataframe
```{r}
sampleNames(flowData) <- SampleSheet$Strain

pData(flowData)$Well <- SampleSheet$Well
pData(flowData)$Strain <- SampleSheet$Strain
pData(flowData)$Genotype <- SampleSheet$Genotype
pData(flowData)$Ploidy <- SampleSheet$Ploidy
pData(flowData)$Media <- SampleSheet$Media
pData(flowData)

```
## View a summary of the data

```{r}
fsApply(flowData,each_col,median)
```

## Visualization using the `ggcyto` package

### Using Autoplot

`autoplot()` is a simple function that provides quick plots of the data with minimal control on the appearance of the plot

#### Ploting a single channel

```{r}
autoplot(flowData, "FL1.A")

```

#### Plotting two channels

```{r}
autoplot(flowData, "FSC.A", "FL1.A")
```

### Using the `ggcyto` function

`ggcyto` provides more control over the plot


#### Ploting a single channel
```{r}
p <- ggcyto(flowData, aes(x = FL1.A))
p + geom_density()
   #    + geom_hex(bins = 128) 
      # + geom_gate("CD8")
```
#### Using `ggridges`

```{r}
p + geom_density_ridges(aes(y = Strain)) + facet_null() #facet_null is used to remove the default facet_wrap (by 'name' column)
```

#### 2D Scatter plots

```{r}
p <- ggcyto(flowData, aes(x = `FSC.A`, y =  `FL1.A`))
p + geom_hex(bins = 128)

```



