---
title: "Gate FACS data"
author: "Gresham Lab"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_notebook:
    toc: yes
    fig_caption: yes
    number_sections: yes
  html_document:
    toc: yes
    df_print: paged
---

This Vignette written as an [R Markdown](http://rmarkdown.rstudio.com) Notebook provides an example of gating FACS data.

The data used in this vignette were obtained from the three FACS machines used in NYU CGSB gencore

1. Accuri
2. Aria
3. Cytek

# Example Data
Below are the data that are used in this vignette.  To work through this analysis:  

1. Download all directories and files from [here](https://drive.google.com/drive/folders/1_QARsH5vOHR-lIPURf3lgt6skN2of5US?usp=sharing)
2. Create and organize a directory `../Data/flowdata` where the data for this vignette reside.

The dataset comprises three different directories with data from three different machines:

* `/Accuri` 
* `/Aria`
* `/Cytek`

each directory contains 

* `.fcs` files
* a samplesheet

```{r}
list.files("/Users/david/Projects/data/flowdata", recursive = T) 
```

# Load required libraries

Several packages are required for running this analysis.  If they are not installed, you need to run the following commands to install them the first time you run this vignette:

## Install Libraries
This workflow uses a variety of Bioconductor packages that should be loaded using: `BiocManager::install()` as per these [guidelines](https://www.bioconductor.org/install/).  It also requires tidyverse, devtools, and ggridges.

```{r Install libraries, warning=FALSE}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#      install.packages("BiocManager")
# BiocManager::install()
# BiocManager::install("ncdfFlow")
# BiocManager::install("flowCore")
# BiocManager::install("flowViz")
# BiocManager::install("ggcyto")
# BiocManager::install("ggforce")
# BiocManager::install("flowWorkspace")
# BiocManager::install("openCyto")
# BiocManager::install("flowStats")
# 
# install.packages("tidyverse")
# install.packages("ggridges")
# install.packages("devtools")
```

## Load the Gresham Lab's `flowtools` from [github](https://github.com/GreshamLab/flowtools)

We have written a set of functions for inhouse flow cytometry analysis that are available in the package `flowtools`

```{r}
library(devtools)
# devtools::install_github("greshamlab/flowtools")
```

## Load libraries
This workflow using the following packages, which must be loaded to run this vignette:

* [flowCore](https://www.bioconductor.org/packages/release/bioc/vignettes/flowCore/inst/doc/HowTo-flowCore.pdf), which provides several functions for reading and manipulating .fcs files.
* [ncdfFlow](https://bioconductor.org/packages/devel/bioc/vignettes/ncdfFlow/inst/doc/ncdfFlow.pdf), which reads .fcs files to disk rather than memory.  This allows import of larger datasets as they do not need to be stored in memory.
* [ggcyto](https://www.bioconductor.org/packages/release/bioc/html/ggcyto.html)
    + [basic functions](https://www.bioconductor.org/packages/release/bioc/vignettes/ggcyto/inst/doc/Top_features_of_ggcyto.html)
    + [autoplot](https://www.bioconductor.org/packages/release/bioc/vignettes/ggcyto/inst/doc/autoplot.html)
    + [visualize flowset](https://www.bioconductor.org/packages/release/bioc/vignettes/ggcyto/inst/doc/ggcyto.flowSet.html)
    + [visualize gatingset](https://www.bioconductor.org/packages/release/bioc/vignettes/ggcyto/inst/doc/ggcyto.GatingSet.html)

```{r}
library(ncdfFlow)
library(flowCore)
library(flowViz)
library(ggcyto)
library(ggforce)
library(tidyverse)
library(ggridges)
library(readxl)
library(flowWorkspace)
library(openCyto)
library(flowStats)
```

# Working with Accuri Data
First we will visualize flow data from the Accuri flow cytometer.

## Read files
FCS files exported from the Accuri flow cytometry machine are read into R using `read.ncdfFlowSet`
```{r}

#read.flow.data() <- function{}

dir = '/Users/david/Projects/data/flowdata/Accuri'
files <- sort(list.files(path=dir,pattern = ".fcs", full.names=TRUE))
flowData <- read.ncdfFlowSet(files=files, pattern=".fcs", alter.names = TRUE)

#needs to differ depending on sample sheet file typ
#sample.sheet <- read_excel(paste(path=dir,"GAP1_multicheck_20May2019.xlsx", sep="/"))
SampleSheet <- read_csv("~/Projects/Data/flowdata/Accuri/SampleSheet.csv")

flowData
```

## Assign sample names and other metadata
Information about the flowset is stored in an __AnnotatedDataFrame__.
The dataframe
```{r}
#annotate.flow.data <- function{}

sampleNames(flowData) <- SampleSheet$Strain

pData(flowData)$Well <- SampleSheet$Well
pData(flowData)$Strain <- SampleSheet$Strain
pData(flowData)$Genotype <- SampleSheet$Genotype
pData(flowData)$Ploidy <- SampleSheet$Ploidy
pData(flowData)$Media <- SampleSheet$Media
pData(flowData)

```
## View a summary of the data

```{r}
fsApply(flowData,each_col,median)

#data_summary <- as_tibble(fsApply(flowData,each_col,median))
#names(data_summary)
```

## Constructing a gate

To perform gating, we use the [openCyto](http://opencyto.org/) package

### Gating doublets and small cells

#### Gating for singlets

```{r}
frame <- flowData[[1]]
sg <- gate_singlet(frame, area = "FSC.A", height = "FSC.H")
autoplot(frame, "FSC.A", "FSC.H", filter = sg)
```



## Visualization using the `ggcyto` package

### Using Autoplot

`autoplot()` is a simple function that provides quick plots of the data with minimal control on the appearance of the plot

#### Ploting a single channel

```{r}
autoplot(flowData, "FL1.A") 

```

#### Plotting two channels

```{r}
autoplot(flowData, "FSC.A", "FL1.A")
```

### Using the `ggcyto` function

`ggcyto` provides more control over the plot


#### Ploting a single channel
```{r}
p <- ggcyto(flowData, aes(x = FL1.A))
p + geom_histogram() + scale_x_flowjo_biexp()
```
#### Using `ggridges`

```{r}
p + scale_x_logicle() + geom_density_ridges(aes(y = Strain)) + facet_null() #facet_null is used to remove the default facet_wrap (by 'name' column)
```

#### 2D Scatter plots

```{r}
p <- ggcyto(flowData, aes(x = `FSC.A`, y =  `FL1.A`))
p + geom_hex(bins = 128)

```

### Using ggplot


```{r}
ggplot(data = flowData, aes(x = FL1.A/FSC.A)) +
  geom_histogram()
```

# Working with Cytek Data
Finally, we will visualize flow data from the Cytek flow cytometer.

## Read files
FCS files exported from the Cytek flow cytometry machine are read into R using `read.ncdfFlowSet`
```{r}
dir = '/Users/david/Projects/data/flowdata/Cytek'
files <- sort(list.files(path=dir,pattern = ".fcs", full.names=TRUE))
CytekflowData <- read.ncdfFlowSet(files=files, pattern=".fcs", alter.names = TRUE)

#needs to differ depending on sample sheet file typ
#sample.sheet <- read_excel(paste(path=dir,"GAP1_multicheck_20May2019.xlsx", sep="/"))
CytekSampleSheet <- read_csv("~/Projects/Data/flowdata/Cytek/SampleSheet.csv")

CytekflowData
```


## Assign sample names and other metadata
Information about the flowset is stored in an __AnnotatedDataFrame__.
The dataframe
```{r}
sampleNames(CytekflowData) <- CytekSampleSheet$Strain

pData(CytekflowData)$Well <- CytekSampleSheet$Well
pData(CytekflowData)$Strain <- CytekSampleSheet$Strain
pData(CytekflowData)$mCitrine <- CytekSampleSheet$mCitrine_copies
pData(CytekflowData)$mCherry <- CytekSampleSheet$mCherry_copies
```


#### Plotting two channels

```{r}
autoplot(CytekflowData, "FSC.A", "V1.A")
```


## View a summary of the data

```{r}
fsApply(CytekflowData,each_col,median)

#data_summary <- as_tibble(fsApply(flowData,each_col,median))
#names(data_summary)
```
